Distributional Trees and Forests for Circular Data

### 2019-10-03 pkg/circtree/inst/wind_paper/draft/draft.Rnw
further revision, particularly on introduction, reference models and motivational example in the application section; added further TODOs
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2784 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 0002dfac5c8c9b2ebc8d2514568fb92c067fd839

Circular data can be found in a variety of applications and subject areas, 
e.g., hourly crime rate in the social-economics, animal movement direction or 
gene-structure in biology, and wind direction as one of the most important weather 
variables in meteorology.
Circular regression models were first introduced by \cite{Fisher+Lee:1992} and 
further extended by \cite{Jammalamadaka+Sengupta:2001} and 
\cite{Mulder+Klugkist:2017} among others.
While most of the already existing approaches are built on additive regression models, 
we propose an adaption of regression trees to circular data by employing 
distributional trees.

@@

\section{Regression Trees and Forests for Circular Data}

\subsection{Motivation}
A commonly used approach to including covariates as regressors in distributional models is 
provided by generalized additive models for location, scale, and shape 
\citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005}.
There, each distribution parameter is linked to a linear predictor by an additive function.
While this strategy works well for smooth effects difficulties can arise in case of 
non-additive changes. Moreover, specifying a model properly can be very challenging,
particularly for a high number of covariates and no information about possible interactions.
An alternative approach that can deal with these problems is to apply a tree-structured model.

@@

\section{Wind Direction Forecasts}

\subsection{Motivation}
Wind is a classical circular quantity and accurate forecasts of wind direction
are of great importance for decision-making processes and risk management,
e.g., in air traffic management or renewable energy production.

@@

This study employs the above introduced methodology of circular distributional
trees and forests to obtain hourly wind direction forecasts at Innsbruck Airport.
Innsbruck lies at the bottom of a deep valley in the Alps. Topography
channels wind along the west-east valley axis or along a tributary valley
intersecting from the south. Hence, pressure gradients to which valley wind
regimes react both west and east of the airport are considered as covariates
along with other meteorological measurements at the airport (lagged by one hour),
such as wind direction and wind speed at Innsbruck Airport.
Note that in the meteorological context wind direction is defined on the scale $[0,360]$ degree and increases clockwise from North ($0$ degree).

@@

Figure~\ref{fig:tree_ibk} depicts the resulting distributional tree, including
both the empirical (gray) and fitted von Mises (red) distribution of wind direction
in each terminal node. Based on the fitted location parameters $\hat \mu$, the subgroups
can be distinguished into the following wind regimes:
(1)~Up-valley winds blowing from the valley mouth towards the upper valley
(from east to west, nodes 4 and 5). (2) Downslope winds blowing across
the Alpine crest along the intersecting valley towards Innsbruck (from south-east to
north-west, nodes 7 and 8). (3) Down-valley winds blowing in the direction
of the valley mouth (from west to east, nodes 12, 14, and 15). Node~11
captures observations with rather low wind speeds that cannot be distinguished
clearly into wind regimes and consequently are associated with a very low
estimated concentration $\hat \kappa$. In terms of covariates, the lagged
wind direction (``persistence'') is mostly responsible for distinguishing
the broad wind regimes listed above while the pressure gradients and wind
speed separate between subgroups with high vs.\ low precision.

\subsection{Data}
Observations of various meteorological quantities from the airport of Innsbruck and .. surrounding
weather stations are used as covariates in order to predict the wind direction at Innsbruck
airport in the following hour.
In particular .. covariates are included in the model. Next to the current wind direction,
temperature, air pressure, ... at the each of the stations also differences in these quantities
between the considered stations are included. For example, differences in air pressure
between Innsbruck and either a western or eastern location are expected to have a high influence 
on wind directions.

\subsection{Models and Evaluation}
Next to distributional trees for circular data and forests three other models have been applied and their 
performances have been compared based on a circular version of the continuous ranked probability 
score (CRPS).

\paragraph{Persistence}
Assuming that weather conditions will remain the same in the following hours/days is the
basic idea of persistence in meteorology. Hence, fitting a distributional model as presented in
Section~\ref{sec:prob_circ} to the observations of the response variable from the previous hour(s) 
contained in the learning data provides a distributional approach to persistence.

\paragraph{Climatology}
If there are no climatological changes it can be expected that for a selected day of the year weather conditions will always be the same. Therefore, in order to predict wind direction for a specified day
and time observations for this selected timespace from all years contained in the learning data are
used to fit a distributional model.

\paragraph{Linear Model}
Another commonly applied model ...

Compare fitted models based on a circular version of CRPS ...

>>>>

Circular data can be found in a variety of applications and subject areas, 
e.g., hourly crime rate in the social-economics, animal movement direction or 
gene-structure in biology, and wind direction as one of the most important weather 
variables in meteorology. Fitting a statistical model to this type of data requires
the incorporation of its specific feature of periodicity, i.e., in case of angular
data the fact that values are restricted to an interval such as $[0,2\pi)$ with
$0$ being equivalent to $2\pi$.

In many approaches to model circular data it is assumed that the circular variable of interest 
follows a circular distribution, in particular the von Mises distribution (which will be discussed 
in more detail in Section~\ref{sec:prob_circ}). This was also the basis on which early regression
models for circular data were built on as they regress the distribution parameter(s) on covariates.
One of the first regression models with a circular response variable and linear covariates
was presented by \cite{Gould:1969} where a linear combination of the covariates is employed to model
the circular mean, hence the location parameter of the von Mises distribtuion. 
\cite{Johnson+Wehrly:1978} took this idea one step further 
by plugging in a link function such that the values of the linear predictor are transformed 
to an restricted interval of length $2\pi$. 
This model was further extended by \cite{Fisher+Lee:1992} and subsequently by \cite{Fisher:1993} 
allowing for the mean direction or the dispersion to be modeled not only in separate models 
but also in a mixed model. However, applying these models has shown to be very challenging, 
mainly due to the complexity encountered in optimizing the log-likelihood function.
Moreover the named approaches and also its further developments 
\citep[see, e.g.,][]{Jammalamadaka+Sengupta:2001, Mulder+Klugkist:2017}
are built on additive regression models which may not be able to capture all types of effects.
For example abrupt changes might not be detected properly owing to the fixed structure enforced
by model assumptions such as the type of link function and the type of linear combination.
Therefore, we propose an adaption of regression trees and random forests to circular data by employing 
distributional trees and distributional forests. 
In that way no model assumptions are required as covariates and their possible interactions are
selected automatically. Non-additive effects are captured by the tree structure while combining an ensemble
of trees to a forest model allows for modeling smooth changes as well. 
Therefore, a high level of flexibility is provided while a full distributional model is specified by distributional trees and forests such that a wide range of inference is made available.
These features and the corresponding methodology are presented in Section~\ref{sec:tree_forest} after discussing 
distributional models for circular data in Section~\ref{sec:prob_circ}.
In Section~\ref{sec:wind} the presented models a applied to probabilistic forecasting of wind direction
at the airports Innsbruck and Vienna.

@@

\section{Regression Trees and Forests for Circular Data}
\label{sec:tree_forest}
\subsection{Motivation}
A commonly used approach to including covariates as regressors in distributional models is 
provided by generalized additive models for location, scale, and shape 
\citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005}.
There, each distribution parameter is linked to a linear predictor by an additive function.
While this strategy works well for smooth effects difficulties can arise in case of 
non-additive changes. Moreover, specifying a model properly can be very challenging,
particularly for a high number of covariates and no information about possible interactions.
An alternative approach that can deal with these problems is to apply a tree-structured model.

@@

\section{Wind Direction Forecasts}
\label{sec:wind}
\subsection{Motivation}
Wind is a classical circular quantity and accurate forecasts of wind direction
are of great importance for decision-making processes and risk management,
e.g., in air traffic management or renewable energy production.

@@

As a motivational example Figure~\ref{fig:tree_ibk} illustrates a distributional
tree applied to predict the wind direction at the airport of Innsbruck in the following hour. 
Innsbruck lies at the bottom of a deep valley in the Alps. Topography channels wind along the 
west-east valley axis or along a tributary valley intersecting from the south. Hence, pressure 
gradients to which valley wind regimes react both west and east of the airport are considered 
as covariates along with other meteorological measurements at the airport (lagged by one hour),
such as wind direction and wind speed at Innsbruck Airport.

Next to the tree structure itself Figure~\ref{fig:tree_ibk} also depicts the empirical (gray) 
and fitted von Mises (red) distribution of wind direction in each terminal node. Based on the 
fitted location parameters $\hat \mu$, the subgroups can be distinguished into the following 
wind regimes:
(1)~Up-valley winds blowing from the valley mouth towards the upper valley (from
east to west, nodes 4 and 5). 
(2) Downslope winds blowing across the Alpine crest along the intersecting valley towards 
Innsbruck (from south-east to north-west, nodes 7 and 8). 
(3) Down-valley winds blowing in the direction of the valley mouth (from west to east, 
nodes 12, 14 and 15). 
Node~11 captures observations with rather low wind speeds that cannot be distinguished
clearly into wind regimes and consequently are associated with a very low
estimated concentration $\hat \kappa$. In terms of covariates, the lagged
wind direction (``persistence'') is mostly responsible for distinguishing
the broad wind regimes listed above while the pressure gradients and wind
speed separate between subgroups with high vs.\ low precision.
Note that in the meteorological context wind direction is defined on the scale $[0,360]$ degree and 
increases clockwise from North ($0$ degree).

@@

Next to the airport of Innsbruck, located in the west of Austria, in the center of the Alps,
a second airport in Austria, the airport of Vienna, situated in the east of the country in a rather flat
area, is considered in this study.
In that way completely different geographic features and their possible influences
on wind direction are included which allows to test for a high flexibility of the applied methods.
For the two locations the above introduced methodology of circular distributional trees and forests is
employed to wind direction data and additional meteorological measurements in order to obtain hourly 
wind direction forecasts, separately for Innsbruck and Vienna.

\subsection{Data}
For both airports observations of various meteorological quantities measured
either directly at the airports or at surrounding weather stations are used as 
covariates in order to predict the wind direction at the corresponding airport 
in the following hour.
Next to the current wind direction, temperature, air pressure, humidity, ... at each 
of the weather stations also differences in these quantities between the considered 
stations are included. For example, differences in air pressure between Innsbruck and 
either a western or eastern location are expected to have a high influence on wind 
directions. Moreover, changes of these quantities over time are included as well as 
daytime. In total, 550 covariates are provided for Innsbruck and 522 covariates for Vienna. 
A list of all covariates together with their abbreviations as used in this paper
can be found in Table~\ref{tab:covariates}.
The data sets each provide observations from 5 years, in particular from 2014 to 2017, both included.
After eliminating observations with at least one missing value a total number of 26861 time points
for Innsbruck and 33853 for Vienna are available.

\begin{table}[t!]
%\resizebox{1.4\textwidth}{!}{
\begin{minipage}{\textwidth}
\caption[Table caption text]{Basic covariates together with the type of variations.}
\label{tab:covariates}
\begin{tabular}{l  l}
\hline
Basic covariates & Variations\\
\hline
\emph{dd}: wind direction       & different locations, \\
                                & changes over time(?)\\
\emph{ff}: wind speed           & different locations, \\
                                & changes over time(?), \\
                                & (hourly) maxima\\
\emph{p}: pressure              & different locations, \\
                                & differences between locations, \\
                                & changes over time(?)\\
\emph{t}: temperature           & different locations, \\
                                & changes over time(?)\\
\emph{rf}: relative humidity    & different locations\\
\hline
\end{tabular}
\end{minipage} 
%}
\end{table}

\subsection{Models and Evaluation}
Along with distributional trees for circular data and forests three other models have been applied 
and their performances have been compared based on a circular version of the continuous ranked probability 
score (CRPS).

\paragraph{Persistence}
Assuming that weather conditions will remain the same after a certain period of time is the
basic idea of persistence in meteorology. While this simple way of making predictions can
be very effective particularly under stable atmospheric circumstances it can fail remarkably
as soon as any changes take place, especially in case of rapid changes.
Therefore, it heavily depends on the specific features and the variability of the variable
of interest whether this approach can provide overall reliable predictions.
% TODO: references
To establish a distributional model of persistence for this study a set of six observations
is considered to make a prediction for a selected day and time point. In particular,
the exact same time points of the previous six days are considered. In order to account
for a stronger influence of observations closer to the prediction time point exponentially
increasing weights are included with the highest weight being assigned to the time point of the
previous day. For this selected set of observations the distribution parameters of the von Mises 
distribution can be estimated as presented in Section~\ref{sec:prob_circ} such that a full
parametric model is specified allowing for probabilistic predictions and an evaluation of the CRPS.

\paragraph{Climatology}
If there are no climatological changes within a certain number of years it can be expected that 
for a selected day of the year weather conditions will always be the same over all years 
considered.
Again, this idea provides a very simple and stable way of obtaining predictions for a considered 
meteorological variable of interest that can be very useful but its performance can also suffer 
dramatically from a high variance and abrupt changes, especially if they occur frequently during one day.
% TODO: references
Similar to the applied persistence model observations are selected to fit a distributional climatology
model. For a specified day and time point observations from the exact same time point and day but all 
other years contained in the learning data are selected and parameters are estimated following again Section~\ref{sec:prob_circ}.

\paragraph{Generalized Linear Model}
Another commonly applied approach is to regress the parameter(s) of the von Mises distribution
on covariates by a linear combination of selected covariates connected to the parameter(s) via
a specified link function. The toolbox of generalized linear models ...
% TODO: references

% TODO: CRPS
\vspace{0.5cm}
Compare fitted models based on a circular version of CRPS ...
